Lattices of acyclic pipe dreams

### 2023-01-20 acyclicFacets/draft/latticePipeDreams.tex
improve intro

\title{The lattice of acyclic pipe dreams}

@@

% theorems
\newtheorem*{theorem*}{Theorem}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem*{conjecture*}{Conjecture}
\newtheorem{conjecture}[theorem]{Conjecture}
\crefname{conjecture}{Conjecture}{Conjectures}

@@

\begin{theorem*}
For any permutation~$\omega$, the Hasse diagram of the lattice quotient of the interval~$[e,\omega]$ of the weak order by the pipe dream congruence of~$\omega$ is isomorphic to the increasing flip graph on acyclic pipe dreams with exiting permutation~$\omega$.
\end{theorem*}

@@

\begin{theorem*}
For any non-empty subword complex~$\subwordComplex(Q, \omega)$, the sets~$\linearExtensions(I)$ for all facets~$I$ of~$\subwordComplex(Q, \omega)$ are order convex and form a partition of a lower set of the weak order that contains the interval~$[e, \omega]$.
\end{theorem*}

@@

\begin{theorem*}
If~$Q$ is sorting, then the sets~$\linearExtensions(I)$ for all facets~$I$ of~$\subwordComplex(Q, \omega)$ form a partition of the interval~$[e, \omega]$.
\end{theorem*}

@@

\begin{conjecture*}
If~$Q$ is alternating, then the subword complex congruence~$\equiv_{Q, \omega}$ is a lattice congruence of the interval~$[e, \omega]$ of the weak order.
\end{conjecture*}

In contrast to the case of pipe dreams, the Hasse diagram of the quotient~$[e, \omega]/{\equiv_{Q, \omega}}$ is not isomorphic to the increasing flip graph on acyclic facets of~$\subwordComplex(Q, \omega)$ for two reasons:
\vincent{refer to the examples...}
\begin{itemize}
\item First, not all acyclic facets of~$\subwordComplex(Q, \omega)$ appear as vertices of~$[e, \omega]/{\equiv_{Q, \omega}}$. We say that a facet is strongly acyclic if~$[e, \omega] \cap \linearExtensions(\omega) \ne \varnothing$.
\item Second, not all flips between two strongly acyclic facets of~$\subwordComplex(Q, \omega)$ define a cover relation of~$[e, \omega]/{\equiv_{Q, \omega}}$. We say that the flip of a position~$i$ in a facet~$I$ is external if the root~$\rootFunction{I}{i}$ directing the flip is a ray of the root configuration~$\Roots(I)$.
\end{itemize}
This leads us to the following conjecture.

\begin{conjecture*}
If~$Q$ is alternating, then the Hasse diagram of the lattice quotient~$[e, \omega]/\equiv_{Q, \omega}$ is isomorphic to the graph of extremal increasing flips between strongly acyclic facets of~$\subwordAcyclicFacets(Q, \omega)$.
\end{conjecture*}

@@

\begin{conjecture*}
If~$Q$ is sorting and alternating, then the Hasse diagram of the lattice quotient~$[e, \omega]/\equiv_{Q, \omega}$ is isomorphic to the graph of extremal increasing flips between acyclic facets of~$\subwordAcyclicFacets(Q, \omega)$.
\end{conjecture*}

This last conjecture has a close connection to the brick polytopes of~\cite{PilaudSanto-brickPolytope, PilaudStump-brickPolytope} and the brick polyhedra of~\cite{JahnStump}.
\vincent{todo}

@@

The \defn{contact graph} of a pipe dream~$P$ is the directed graph~$P\contact$ with one node for each pipe of~$P$ and one arc for each contact of~$P$ connecting the northwest pipe to the southeast pipe of the contact\footnote{We have reversed the usual orientation conventions of \cite{PilaudSantos-brickPolytope, PilaudPocchiola, Pilaud-BrickAlgebra} to suit better our purposes, in particular in \cref{subsec:insertionAlgorithm}.}.
We see equivalently the contact graph~$P\contact$ as a graph on the pipes of~$P$ or on the integers~$[n]$.
We say that a pipe dream~$P$ is \defn{acyclic} if its contact graph~$P\contact$ is acyclic (no oriented cycle).
We then denote by~$\contactLess{P}$ the transitive closure of the contact graph of~$P$.
For~$\omega \in \fS_n$, we denote by~$\acyclicPipeDreams(\omega)$ the set of acyclic pipe dreams of~$\pipeDreams(\omega)$.
% We let~$\acyclicPipeDreams_n \eqdef \bigsqcup_{\omega \in \fS_n} \acyclicPipeDreams(\omega)$ and~$\acyclicPipeDreams \eqdef \bigsqcup_{n \in \N} \acyclicPipeDreams_n$.
%\vincent{Add contact graphs to \cref{fig:pipeDreams}.}

@@

Our second algorithm to construct~$\insertion{\pi}{\omega}$ is an \defn{insertion algorithm} inspired from~\cite{Pilaud-BrickAlgebra} and similar to the insertion in binary search trees.

@@

\begin{example}\label{ex_acyclicflip_not_weakorder_related}
Consider the type~$A_2$ Coxeter system, the word~$Q = \tau_1 \tau_2 \tau_1 \tau_2 \tau_1 \tau_2$ and the longest element~$\wo = \tau_1 \tau_2 \tau_1 = \tau_2 \tau_1 \tau_2$.

@@

\begin{remark}\label{rem_acyclicflipposet_not_latticequotient}
In analogy to the case of pipe dreams in \cref{subsec:proofCongruence}, one might expect that the quotient~$[e,\omega]/\equiv_{Q,\omega}$ is isomorphic to the restriction of the increasing flip poset of $\subwordComplex(Q,\omega)$ to the set of acyclic facets.
But this is not the case as shown in \cref{ex_acyclicflip_not_weakorder_related}.     
\end{remark}

\begin{example}\label{ex_acyclicflip_not_weakorder_related}
    Let $W=S_3$ be the symmetric group generated by simple transpositions $S=\{s_1,s_2\}$ with $s_i=(i, i+1 )$.
Let $Q=(s_1,s_2,s_1,s_2,s_1,s_2)$ and $\omega=\wo=s_1s_2s_1=s_2s_1s_2$.
The subword complex $\subwordComplex(Q,\omega)$ has eight facets, six of which are acyclic.
    All the fibers of $\equiv_{Q,\omega}$ are singletons and the lattice quotient~$[e,\omega]/\equiv_{Q,\omega}$ coincides with the weak order of type $A_2$.
    On the other hand there are two facets, $I_1=\{1,3,4\}$ and $I_2=\{3,4,6\}$, which are connected by a flip but whose corresponding elements $\pi_1$ and $\pi_2$ are not related in weak order.
    \cesar{To do: Double check and expand this example in detail. Add figure.}
\end{example}

@@

As pointed out in \cref{rem_acyclicflipposet_not_latticequotient},  we can not expect~$[e,\omega]/\equiv_{Q,\omega}$ to be isomorphic to the restriction of the increasing flip poset of $\subwordComplex(Q,\omega)$ to the set of acyclic facets.
\cref{ex_acyclicflip_not_weakorder_related} shows that there are flips between acyclic facets that are not detected by the week order.
One could expect that there is a natural restriction to a special class of flips that behave nicer, meaning that all of them are identified by the quotient~$[e,\omega]/\equiv_{Q,\omega}$.
In our \cref{ex_acyclicflip_not_weakorder_related}, the generalized Chute moves from~\cite{Rubey} do the job, because the flip that is not detected by the week order is exactly the one flip which is not Chute.
The Chute lattice conjecture~\cite[Conjecture~2.8]{Rubey} might be an indication that this is the right direction to look at.

>>>>

\title{Lattices of acyclic pipe dreams}

@@

% theorems
\newtheorem{theorem}{Theorem}[section]
\newtheorem{theoremA}{Theorem}
\renewcommand{\thetheoremA}{\Alph{theoremA}}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{conjectureA}{Conjecture}
\renewcommand{\theconjectureA}{\Alph{conjectureA}}
\crefname{conjecture}{Conjecture}{Conjectures}

@@

\begin{theoremA}
For any permutation~$\omega$, the Hasse diagram of the lattice quotient of the interval~$[e,\omega]$ of the weak order by the pipe dream congruence of~$\omega$ is isomorphic to the increasing flip graph on acyclic pipe dreams with exiting permutation~$\omega$.
\end{theoremA}

@@

\begin{theoremA}
For any non-empty subword complex~$\subwordComplex(Q, \omega)$, the sets~$\linearExtensions(I)$ for all facets~$I$ of~$\subwordComplex(Q, \omega)$ are order convex and form a partition of a lower set of the weak order that contains the interval~$[e, \omega]$.
\end{theoremA}

@@

\begin{theoremA}
If~$Q$ is sorting, then the sets~$\linearExtensions(I)$ for all facets~$I$ of~$\subwordComplex(Q, \omega)$ form a partition of the interval~$[e, \omega]$.
\end{theoremA}

@@

\begin{conjectureA}
If~$Q$ is alternating, then the subword complex congruence~$\equiv_{Q, \omega}$ is a lattice congruence of the interval~$[e, \omega]$ of the weak order.
\end{conjectureA}

In contrast to the case of pipe dreams, the Hasse diagram of the quotient~$[e, \omega]/{\equiv_{Q, \omega}}$ is not isomorphic to the increasing flip graph on acyclic facets of~$\subwordComplex(Q, \omega)$ for two reasons:
\begin{itemize}
\item First, not all acyclic facets of~$\subwordComplex(Q, \omega)$ appear as vertices of~$[e, \omega]/{\equiv_{Q, \omega}}$ (see \cref{rem:linearExtensionsPartitionSubwordComplexA}). We say that a facet is strongly acyclic if~$[e, \omega] \cap \linearExtensions(\omega) \ne \varnothing$.
\item Second, not all flips between two strongly acyclic facets of~$\subwordComplex(Q, \omega)$ define a cover relation of~$[e, \omega]/{\equiv_{Q, \omega}}$ (see \cref{exm:acyclicFlipvsWeakOrder}). We say that the flip of a position~$i$ in a facet~$I$ is external if the root~$\rootFunction{I}{i}$ directing the flip is a ray of the root configuration~$\Roots(I)$.
\end{itemize}
This leads us to the following conjecture.

\begin{conjectureA}
If~$Q$ is alternating, then the Hasse diagram of the quotient~$[e, \omega]/\equiv_{Q, \omega}$ is isomorphic to the graph of extremal increasing flips between strongly acyclic facets of~$\subwordAcyclicFacets(Q, \omega)$.
\end{conjectureA}

@@

\begin{conjectureA}
If~$Q$ is sorting and alternating, then the Hasse diagram of the quotient~$[e, \omega]/\equiv_{Q, \omega}$ is isomorphic to the graph of extremal increasing flips between acyclic facets of~$\subwordAcyclicFacets(Q, \omega)$.
\end{conjectureA}

This last conjecture has a strong connection to the geometry of subword complexes given by brick polytopes of~\cite{PilaudSantos-brickPolytope, PilaudStump-brickPolytope} and the brick polyhedra of~\cite{JahnStump}.
Namely, the extremal increasing flips between acyclic facets of~$\subwordAcyclicFacets(Q, \omega)$ is precisely the bounded part (meaning forgetting the unbounded rays) of the brick polyhedron~$\brickPolyhedron(Q, \omega)$ oriented by a natural linear functional.
In particular, when~$\omega = \wo$, we obtain the following conjecture, extending results from~\cite{Pilaud-brickAlgebra}.

\begin{conjectureA}
If~$Q$ is sorting and alternating, the oriented graph of the brick polytope~$\brickPolyhedron(Q, \wo)$ is the Hasse diagram of a lattice quotient of the weak order.
\end{conjectureA}

@@

The \defn{contact graph} of a pipe dream~$P$ is the directed graph~$P\contact$ with one node for each pipe of~$P$ and one arc for each contact of~$P$ connecting the northwest pipe to the southeast pipe of the contact\footnote{We have reversed the usual orientation conventions of \cite{PilaudSantos-brickPolytope, PilaudPocchiola, Pilaud-brickAlgebra} to suit better our purposes, in particular in \cref{subsec:insertionAlgorithm}.}.
We see equivalently the contact graph~$P\contact$ as a graph on the pipes of~$P$ or on the integers~$[n]$.
We say that a pipe dream~$P$ is \defn{acyclic} if its contact graph~$P\contact$ is acyclic (no oriented cycle).
We then denote by~$\contactLess{P}$ the transitive closure of the contact graph of~$P$.
For~$\omega \in \fS_n$, we denote by~$\acyclicPipeDreams(\omega)$ the set of acyclic pipe dreams of~$\pipeDreams(\omega)$.
% We let~$\acyclicPipeDreams_n \eqdef \bigsqcup_{\omega \in \fS_n} \acyclicPipeDreams(\omega)$ and~$\acyclicPipeDreams \eqdef \bigsqcup_{n \in \N} \acyclicPipeDreams_n$.
%\vincent{Add contact graphs to \cref{fig:pipeDreams}.}

@@

Our second algorithm to construct~$\insertion{\pi}{\omega}$ is an \defn{insertion algorithm} inspired from~\cite{Pilaud-brickAlgebra} and similar to the insertion in binary search trees.

@@

\begin{example}
\label{exm:acyclicFlipvsWeakOrder}
Consider the type~$A_2$ Coxeter system, the word~$Q = \tau_1 \tau_2 \tau_1 \tau_2 \tau_1 \tau_2$ and the longest element~$\wo = \tau_1 \tau_2 \tau_1 = \tau_2 \tau_1 \tau_2$.

@@

\begin{remark}\label{rem_acyclicflipposet_not_latticequotient}
In analogy to the case of pipe dreams in \cref{subsec:proofCongruence}, one might expect that the quotient~$[e,\omega]/\equiv_{Q,\omega}$ is isomorphic to the restriction of the increasing flip poset of $\subwordComplex(Q,\omega)$ to the set of acyclic facets.
But this is not the case as shown in \cref{exm:acyclicFlipvsWeakOrder}.     
\end{remark}

\begin{example}\label{exm:acyclicFlipvsWeakOrder}
    Let $W=S_3$ be the symmetric group generated by simple transpositions $S=\{s_1,s_2\}$ with $s_i=(i, i+1 )$.
Let $Q=(s_1,s_2,s_1,s_2,s_1,s_2)$ and $\omega=\wo=s_1s_2s_1=s_2s_1s_2$.
The subword complex $\subwordComplex(Q,\omega)$ has eight facets, six of which are acyclic.
    All the fibers of $\equiv_{Q,\omega}$ are singletons and the lattice quotient~$[e,\omega]/\equiv_{Q,\omega}$ coincides with the weak order of type $A_2$.
    On the other hand there are two facets, $I_1=\{1,3,4\}$ and $I_2=\{3,4,6\}$, which are connected by a flip but whose corresponding elements $\pi_1$ and $\pi_2$ are not related in weak order.
    \cesar{To do: Double check and expand this example in detail. Add figure.}
\end{example}

@@

As pointed out in \cref{rem_acyclicflipposet_not_latticequotient},  we can not expect~$[e,\omega]/\equiv_{Q,\omega}$ to be isomorphic to the restriction of the increasing flip poset of $\subwordComplex(Q,\omega)$ to the set of acyclic facets.
\cref{exm:acyclicFlipvsWeakOrder} shows that there are flips between acyclic facets that are not detected by the week order.
One could expect that there is a natural restriction to a special class of flips that behave nicer, meaning that all of them are identified by the quotient~$[e,\omega]/\equiv_{Q,\omega}$.
In our \cref{exm:acyclicFlipvsWeakOrder}, the generalized Chute moves from~\cite{Rubey} do the job, because the flip that is not detected by the week order is exactly the one flip which is not Chute.
The Chute lattice conjecture~\cite[Conjecture~2.8]{Rubey} might be an indication that this is the right direction to look at.
